---
title: "Lab 9: Data Wrangling with `dplyr`"
author: "STAT 630, Fall 2021"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

The R package `dplyr` provides a set of functions for data manipulation, or data wrangling. `dplyr` is one of the core packages in the so-called tidyverse, which also includes `ggplot2`.  

For this lab, we will focus on the following commonly used `dplyr` functions:

* `select()` take a subset of the columns (variables)

* `filter()` take a subset of the rows (observations)

* `arrange()` reorder the rows

* `mutate()` creates new variables that are functions of existing variables

* `summarise()` and `group_by()` compute summary statistics across different groups

The names of these functions are *verbs* that provide a grammar for data wrangling.\

To load `dplyr` into your R session run the following command:
```{r, message = FALSE}
library(dplyr)
```
&nbsp;

`dplyr` reference: https://dplyr.tidyverse.org/index.html

\newpage

## CDC Data Set

We will use the CDC data set one more time to demonstrate how to use the `dplyr` package.  The variable descriptions are given in lab 2.

```{r}
cdc <- readRDS(url("https://ericwfox.github.io/data/cdc.rds"))
dim(cdc)
head(cdc)
```

## `select()`

Use `select()` to subset the columns (variables) of a data frame.

```{r}
cdc2 <- select(cdc, weight, age, gender)
head(cdc2)
```

```{r, eval=F}
# equivalent base R command
cdc2 <- cdc[, c("weight", "age", "gender")]
```

## filter()

Use `filter()` to subset the rows of a data frame.  The first argument is the name of the data frame.  The second argument is a **logical expression** that specifies the rows to subset.

```{r}
cdc2 <- filter(cdc, age <= 30 & gender == "m")
head(cdc2)
```

```{r, eval=F}
# equivalent base R command
cdc2 <- subset(cdc, age <= 30 & gender == "m")
```


The following table summarizes the different logical operators in R that you can use with `filter()`:

\begin{center}
\begin{tabular}{|r|r|}
\hline
Operator & Description\\
\hline
\texttt{<} & less than\\
\texttt{<=} & less than or equal to\\
\texttt{>} & greater than\\
\texttt{>=} & greater than or equal to\\
\texttt{==} & exactly equal to\\
\texttt{!=} & not equal to\\
\texttt{x | y} & x OR y\\
\texttt{x \& y} & x AND y\\
\hline
\end{tabular}
\end{center}

## `%>%` 

The pipe operator `%>%` can be used combine multiple operations together.

```{r}
cdc %>%
  select(weight, age, gender) %>%
  filter(age <= 30 & gender == "m") %>%
  head()
```
&nbsp;

**Exercise (in-class):**  Use `filter()` to subset the rows of the `cdc` data frame corresponding to individuals that 

1. are over 6 feet tall

2. are men that have smoked over 100 cigarettes

3. are in `good` or `very good` health  

```{r, eval=F, echo=F}
cdc %>% filter(height > 72) %>% head(n=10)
```

```{r, eval=F, echo=F}
cdc %>% filter(smoke100 == 1 & gender == "m")  %>% head(n=10)
```

```{r, eval=F, echo=F}
cdc %>% filter(genhlth == "good" | genhlth == "very good") %>% head(n=10)
```



\newpage

## `arrange()`

Use `arrange()` to order the rows of a data frame by the values of a column.

```{r}
cdc %>% arrange(weight) %>% head(n=10)
```

Use `desc()` to reorder in descending order:
```{r}
cdc %>% arrange(desc(weight)) %>% head(n=10)
```

## `mutate()`

Use `mutate()` to add a new variable to a data frame.

```{r}
cdc %>%
  mutate(wtdiff = wtdesire - weight) %>%
  head(n=10)
```

\newpage

## `group_by()` and `summarise()`

`group_by()` and `summarise()` can be used to compute summary statistics across different groups (categories).

The following code gives the mean weight and height for each category of `gender`.  The counts are also given using the `n()` function in `sumamrise()`.  

```{r, message = FALSE}
cdc %>%
  group_by(gender) %>%
  summarise(
    count = n(),
    weight_mean = mean(weight),
    height_mean = mean(height)
  )
```
&nbsp;

**Exercises (in-class):**  For each category of `genhlth` compute the mean age, weight, and desired weight.

```{r, eval=F, echo=F}
hlth_levels <-  c("poor", "fair", "good", "very good", "excellent")
cdc$genhlth<- factor(cdc$genhlth, levels = hlth_levels)
cdc %>%
  group_by(genhlth) %>%
  summarise(
    count = n(),
    age_mean = mean(age),
    weight_mean = mean(weight),
    wtdesire_mean = mean(wtdesire)
  )
```





