---
title: "Lab 7 Solutions, STAT 630"
output: pdf_document
---

## Exercise 1
### a) Compute summary statistics and plot a histogram and QQ plot for chlorine concentrations.
```{r, fig.width=8, fig.height=4, message=F, warning=F}
library(resampledata)
Chlorine <- Bangladesh$Chlorine
Chlorine <- Chlorine[!is.na(Chlorine)]
summary(Chlorine)
par(mfrow=c(1,2))
hist(Chlorine, main="")
qqnorm(Chlorine)
qqline(Chlorine)
```

### b) Generate 5000 bootstrap replicates of the mean.  Make a histogram and QQ plot of the 5000 replicates that you generated.
```{r}
set.seed(9999)
n <- length(Chlorine)
replicates <- rep(0, 5000)
for(i in 1:5000) {
  boot_samp <- sample(Chlorine, size = n, replace = TRUE)
  replicates[i] <- mean(boot_samp) 
}
```
\clearpage

```{r, fig.width=8, fig.height=4}
par(mfrow=c(1,2))
hist(replicates, xlab="Replicates of the Mean", main='')
abline(v=mean(Chlorine), col="cyan")
qqnorm(replicates)
qqline(replicates)
```


### c) Calculate a 95\% bootstrap percentile confidence interval for the mean.  Interpret the interval.

```{r}
bootCI <- quantile(replicates, c(0.025, 0.975))
round(bootCI, 2)
```
Interpretation: We are 95\% confident that the true mean chlorine concentration is between 54.76 and 104.50 micrograms per liter.

### d) Calculate a traditional  95\% confidence interval for the mean using the $z$-critical value.  Compare this interval with the bootstrap confidence interval you calculated.

```{r}
ci_lower <- mean(Chlorine) - 1.96 * sd(Chlorine) / sqrt(n)
ci_upper <- mean(Chlorine) + 1.96 * sd(Chlorine) / sqrt(n)
round(c(ci_lower, ci_upper), 2)
```

We see that the endpoints of the 95\% bootstrap interval and z-interval are close.  
\clearpage

## Exericse 2
### a) Generate 5000 bootstrap replicates of the median.  Make a histogram and QQ plot of the 5000 replicates that you generated.
```{r}
set.seed(9999)
mean(Chlorine, trim=0.1)
replicates <- rep(0, 5000)
for(i in 1:5000) {
  boot_samp <- sample(Chlorine, size = n, replace = TRUE)
  replicates[i] <- mean(boot_samp, trim=0.1) 
}
```

```{r, fig.width=6, fig.height=2.75}
par(mfrow=c(1,2), cex=0.7)
hist(replicates, breaks=10, xlab="Replicates of the Median", main='')
abline(v=mean(Chlorine, trim=0.1), col="cyan")
qqnorm(replicates); qqline(replicates)
```

### b) Calculate the bootstrap standard error for the median.
```{r}
sd(replicates)
```

### c) Calculate a 95\% bootstrap percentile confidence interval for the median.  Interpret the interval.
```{r}
bootCI <- quantile(replicates, c(0.025, 0.975))
round(bootCI, 2)
```
Interpretation: We are 95\% confident that the true median chlorine concentration is between 11.5 and 16.6 micrograms per liter.

